A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-14 paper.tex
Updates from Overleaf

% Second authors
\author[2,$\star$]{Anastasia Ignatieva}
\author[3,$\star$]{Jere Koskela}

% Middle Authors
\author[4]{Gregor Gorjanc}
\author[5,6]{Anthony W. Wohns}

@@

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
\affil[3]{Department of Statistics, University of Warwick, CV4 7AL, UK}
\affil[4]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
\affil[5]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
\affil[6]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}

@@

% FIXME this is weak, but gives the basic outline
\begin{abstract}
Recent breakthroughs have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an Ancestral Recombination Graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the coalescent with recombination
stochastic process, and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, and that these assumptions are
% Better wording
substantially broken
in modern large-scale datasets.
Representing genetic ancestry
with an eARG also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the detailed genetic genealogy of a set of sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as inputs have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was defined as a variant 
of the coalescent with recombination stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
but has subsequently come to be seen as a data
structure~\citep{minichiello2006mapping}, i.e., a means of 
describing the \emph{outcome} of some such ancestral process
(the distinction, however, is not clear cut and subfields use the term
differently; see Appendix XXX).
ARG inference was for many years synonymous [FIXME: this isn't really true,
what about parsimony ARGs?]
with estimating realisations
of the coalescent with recombination (with, it must be said, limited success),
and the distinction between process and realisation was not especially important.
The recent introduction of two methods, \relate~\citep{speidel2019method}
and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
sharing many properties of this classical idea of an ARG, but at previously
unthinkable scale, has led to a surge of interest in the area.
Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
avoid describing the output of the methods as an ARG, the general 
consensus has emerged that they belong to a growing class of 
ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
and we therefore need a more general definition of an ``ARG'' than the classical
Griffiths formulation.

A general term to describe the detailed genetic genealogy of a sample 
that encompasses the output of all these methods is indeed useful, but 
there are substantial problems emerging from the current usage of the term 
ARG (and its implicit concrete definition in the Griffiths formulation).
The problem is much deeper than some pedantic discussion over terminology,
and directly informs what we should be attempting to estimate, how we 
evaluate inference methods, and what the limitations on the precision
and accuracy of our inferences are.
We currently lack a well-defined target for inference, and therefore there is 
disagreement on how these inference should be evaluated, and 
what the criteria for success should be.
We also lack a shared format for interchanging ARG data between
programs, with the assorted negative consequences which
geneticists are only too familiar with~\citep{excoffier2006computer}.
Finally, and perhaps most importantly, and overly narrow definition of ARGs
leads to potentially missed opportunities in their application and usage.

% What do we do?
In this paper we attempt to address these problems by 
providing a precise definition of an ARG that is general enough to encompass
the output of all currently available inference methods. 
The detailed genetic genealogy of a sample is defined in terms 
of based on specific genomes
and their intervals of genetic inheritance, which we refer to as 
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach which 
is based on recording evolutionary events, and we therefore refer to as 
the ``event ARG'', or ``eARG'' encoding.
We show that the gARG encoding has a range of advantages over the 
eARG approach, being both flexible and efficient.
In particular, we argue that the goal of inferring fully precise 
ARGs---a necessity under the eARG encoding---should be critically 
examined as part of the process of fully decoupling our ARG data 
structure from the coalescent with recombination model.

@@

% % \node [anchor=north west] at (9,6) {C};
% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
% % \begin{tabular}{ll}
% % Edges & Intervals\\
% % \hline
% % $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
% % $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
% % $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
% % $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
% % $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
% % $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
% % $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
% % \end{tabular}};

@@

% Fortunately, there is a straightforward change to the
% classical eARG encoding that simplifies
% implementations and greatly increases the flexibility
% of both the types and granularity of inheritance information
% that we can represent.
% The main change that we need to make is to associate
% the details of inheritance with \emph{edges} rather than
% as breakpoints associated with recombination events and
% edge ordering rules.
% This is illustrated in Fig~\ref{fig-arg-data-structure}C
% where this alternative edge annotation formulation is
% shown. By associating the interval $(0, x)$ with edge $(d,e)$
% and the interval $(x, L)$ with $(d, f)$ we make the encoding
% unambiguous: we can list the edges in any order, and it
% will describe the same set of local genealogical trees.
% This approach is also more flexible, because any pattern
% of inheritance can be described by these intervals. For
% example, if a gene conversion event occurred where a segment
% between positions $x$ and $y$ was inherited by
% node $d$ from $e$, we would state this by associating
% the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
% $\{(x, y]\}$ with $(d,f)$.
% This change is straightforward, involves no loss of information
% and (as we see in later sections)
% greatly increases both the expressivity and computational
% power of ancestral recombination graph data structures.

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of specific genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] which persist over 
many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

A gARG is then a graph in which nodes represent
monoploid genomes and edges represent
genetic inheritance between an ancestor and a descendant,
where each 
edge\footnote{We use the standard graph theoretic sense of
an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
 is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but
where each individual is represented by two nodes (for diploids).
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal  and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are independent recombinants of
the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
% Recombination (e.g.,
% the cause of the breakpoint at position 7)
% is associated with two parent genomes (\textsf{e} and
% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
% (C) The edge-annotated genome ARG in (B) fully defines the
(C) The corresponding local (``marginal'') trees.
}
\end{figure}

The edges in a gARG specify that genetic inheritance occurred between an
ancestor and a descendant, but the graph connectivity 
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome 
from an ancestor) we must ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $(0, 2]$ from \textsf{e} and it inherits $(2, 10]$ from \textsf{f}.
These intervals are shown attached to the corresponding graph edges.

% JK: dunno if this will make some readers switch off, but I feel it's 
% important to show that this is a formal model and the concrete definition
% will help some readers?
Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $(\ell, r]$ over
which genome $c$ inherits from $p$. The most fundamental operation for
an ARG data structure is the process of recovering the corresponding sequence
of local trees along the genome. To generate the local tree at genome
coordinate $x$, we traverse the graph rootwards from the leaves. 
For concreteness, we can describe this tree as an oriented forest $\pi$
\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
At each node
$u$, we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$.
We then set $\pi_c \leftarrow p$,  and move to the next node $p$.
At the end of this process, $\pi$ fully describes the local tree 
at position $x$.

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we record the intervals of inheritance
between genomes (and not causal \emph{events}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple recombinations
across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, ``local'' trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node \noderef{e}
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that \noderef{e} the ancestor of \noderef{a} over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
% TODO rephrase this, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

% This paragraph is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?
% There are some interesting and important consequences to this focus on genomes
% rather than events.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
The final vital
detail is that we associate a \emph{breakpoint} with each recombination
event, and this information is sufficient to uniquely
define the local genealogical trees at every position along the genome.

As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).

>>>>

% Second authors
\author[2,$\star$]{Anastasia Ignatieva}
\author[3,4,$\star$]{Jere Koskela}

% Middle Authors
\author[5]{Gregor Gorjanc}
\author[6,7]{Anthony W. Wohns}

@@

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
\affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
\affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
\affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
\affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
\affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}

@@

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an Ancestral Recombination Graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the coalescent with recombination
stochastic process, and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, and that these assumptions are
% Better wording
substantially broken
in modern large-scale datasets.
Representing genetic ancestry
with an eARG also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the detailed genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics. 
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these ancestries as inputs have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was based on a stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral} which modelled the coalescent with recombination \citep{hudson1983properties}. Subsequently, it has come instead to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ a means of 
describing the sequence of correlated local trees that are the \emph{outcome} of some such ancestral process. However, the distinction between process and structure is not clear cut and subfields use the term
differently (see Appendix XXX).
%% Sentence deleted: point made already
% ARG inference was for many years synonymous [FIXME: this isn't really true,
% what about parsimony ARGs?]
% with estimating realisations
% of the coalescent with recombination (with, it must be said, limited success),
% and the distinction between process and realisation was not especially important.

%% Repeat of first para
% The recent introduction of two methods, \relate~\citep{speidel2019method}
%and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
%sharing many properties of this classical idea of an ARG, but at previously
%unthinkable scale, has led to a surge of interest in the area.
% 
%
%Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
%avoid describing the output of the methods as an ARG, 
%and we therefore need a more general definition of an ``ARG'' than the classical
%Griffiths formulation.

The word ``ARG'' is now widely used to refer to genetic genealogies in general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree, kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, some definitions of ARGs (especially those linked to a specific stochastic process) are overly narrow, which
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we 
assess success, and the expected limitations on the precision
and accuracy of our inferences.

% What do we do?
In this paper we attempt to address these problems by 
providing a precise definition of an ARG that is general enough to encompass
the output of all popular simulation and inference methods, yet simple enough to be encoded via a conventional mathematical graph. 
%% Yan: I added the above as both the ARGON and the Shipilova definitions
%% use non-conventional graph structures for their encoding (e.g. hypergraphs, weird edges, etc)
We follow the well established principle that it is advantageous to decouple the presence of mutations from the genealogical structure itself \citep[contra][chap.~3]{gusfield2014recombinatorics}. We propose a definition of genetic genealogy purely in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as 
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach, 
which is based on recording evolutionary events, and which we call 
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the 
eARG approach, being both flexible and efficient.
In particular, we argue that the goal of inferring fully precise 
ARGs---a necessity under the eARG encoding---should be 
% weird phrasing for a criticism?
critically examined
%
as part of the process of fully decoupling our ARG data 
structure from any specific generative model.

@@

% % \node [anchor=north west] at (9,6) {C};
% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
% % \begin{tabular}{ll}
% % Edges & Intervals\\
% % \hline
% % $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
% % $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
% % $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
% % $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
% % $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
% % $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
% % $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
% % \end{tabular}};

@@

% Fortunately, there is a straightforward change to the
% classical eARG encoding that simplifies
% implementations and greatly increases the flexibility
% of both the types and granularity of inheritance information
% that we can represent.
% The main change that we need to make is to associate
% the details of inheritance with \emph{edges} rather than
% as breakpoints associated with recombination events and
% edge ordering rules.
% This is illustrated in Fig~\ref{fig-arg-data-structure}C
% where this alternative edge annotation formulation is
% shown. By associating the interval $(0, x)$ with edge $(d,e)$
% and the interval $(x, L)$ with $(d, f)$ we make the encoding
% unambiguous: we can list the edges in any order, and it
% will describe the same set of local genealogical trees.
% This approach is also more flexible, because any pattern
% of inheritance can be described by these intervals. For
% example, if a gene conversion event occurred where a segment
% between positions $x$ and $y$ was inherited by
% node $d$ from $e$, we would state this by associating
% the intervals $\{[0, x), [y, L)\}$ with $(d,e)$ and
% $\{[x, y)\}$ with $(d,f)$.
% This change is straightforward, involves no loss of information
% and (as we see in later sections)
% greatly increases both the expressivity and computational
% power of ancestral recombination graph data structures.

\section*{Genome ARGs}

We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] that persist over 
many generations. A ``genome'' in our usage corresponds to the 
haploid genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

A gARG is then a graph in which nodes represent
haploid genomes, edges represent
genetic inheritance between an ancestor and a descendant, and
annotations specify coordinates of the genomic intervals
over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree.
An example of this for diploids (so each individual is represented by two nodes) is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
Each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
% Recombination (e.g.,
% the cause of the breakpoint at position 7)
% is associated with two parent genomes (\textsf{e} and
% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
% (C) The edge-annotated genome ARG in (B) fully defines the
(C) The corresponding local (``marginal'') trees.
}
\end{figure}

The edges in a gARG specify that genetic inheritance occurred between an
ancestor and a descendant, but the graph connectivity 
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome 
from an ancestor) we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f}; these intervals are shown attached to the corresponding graph edges. Alternative gARG encodings split transmitted intervals and associate each with a new node, resulting in a single genome being represented by a group of multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for efficiency of storage.

% JK: dunno if this will make some readers switch off, but I feel it's 
% important to show that this is a formal model and the concrete definition
% will help some readers?
Formally, our gARG encoding is defined by sets of nodes $N$ and edges $E$, where each
element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
which genome $c$ inherits from $p$. The most fundamental operation for
an ARG data structure is the process of recovering the corresponding sequence
of local trees along the genome. Concretely, we can describe a local tree
at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$, where the entry $\pi_j$ is the index
of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the 
tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
We use a forest to describe the local ancestry at a given position so that we allow for cases where not all leaves have found a most recent common ancestor, resulting in multiple roots.
To construct $\pi$ at position $x$, set $\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$. 
Then, traversing over all nodes of the tree,
at each node
$u$, find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$.
At the end of this process, $\pi$ fully describes the local tree.

There are some interesting consequences to these definitions. Firstly,
there is
no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e.\ two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we record the intervals of inheritance
between genomes (and not causal \emph{events}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations (e.g.\ Figure~\ref{fig-arg-in-pedigree}), and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, ``local'' trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node \noderef{e}
is present but not coalescent in the left-most tree of Figure~\ref{fig-arg-in-pedigree} (see the ``Precision of recombination information'' section for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

In this section we have considered the nodes in a gARG as separate from events in a generative process.
Nevertheless, gARGs are commonly constructed to include only those genomes associated with specific events.
For example, \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree... contain[ing] only those edges along which inherited segments of DNA have been transmitted, and only those nodes
corresponding to ancestors in which there was a recombination or coalescence
event.''. In an eARG however, events are even more explicitly embedded in the data structure.

% There are some interesting and important consequences to this focus on genomes
% rather than events.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a backward-time stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an ARG data structure is in terms of events, and we therefore refer to it as an ``event ARG'' (eARG).
In an eARG, nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages, and a list of two parents is required.
As illustrated in Figure~\ref{fig-arg-data-structure},
the regions of the genome which are transmitted are usually
specified by associating recombination nodes with a breakpoint.
In other words, eARGs traditionally
carry positional information on their nodes, rather than
storing genome positions as edge annotations.

To recover the local tree at position $x$,
requires traversing the graph rootwards from the leaves.
When we encounter a recombination node during traversal
if $x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent. This means the order in which parents are
listed is important. Alternatively, additional information can be incorporated to distinguish left and right parents \citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.
